Higher-Order unification for free!

### 2024-04-15 paper/main.tex
minor

\input{code/progress_heuristic}

\noindent
Note that \elpiIn{UV} is in \llambda as well as \elpiIn{T}.

\section{Actual implementation in Elpi}\label{sec:implementation}

@@

\begin{elpicode}
constraint link-eta {
  rule (N1 ~$\triangleright$~ G1 ?- link-eta (uvar X LX1) T1)   % match
    /  (N2 ~$\triangleright$~ G2 ?- link-eta (uvar X LX2) T2)   % remove
    |  (relocate LX1 LX2 T2 T2')               % condition
   <=> (N1 ~$\triangleright$~ G1 ?- T1 = T2').                } % new goal
\end{elpicode}

\noindent
The first directive matches a constraint
whose \lhs is a variable \elpiIn{X}; the second mathes \emph{and removes}
a constrain on the same variable; the third is a test that accessorily
relocates \elpiIn{T2} and the latter crafs the new goal that unifies
\elpiIn{T1} with \elpiIn{T2'} under \elpiIn{G1} and the set of
heigenvariables \elpiIn{N1}.

@@

In the literature we could find a related encoding of the Calculus of
Constructions (CC)~\cite{felty93lics}. The goal of that work is to exhibit
a logic program performing proof checking for CC and hence relate the
proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
programs) with the one of CC. The encoding is hence tailored
toward a different goal, for example it utilizes three relations to represent the
equational theory of CC, and that choice alone makes things arder for us.
Section 6 contains a discussion about the use of the
unification procedure of the ML in presence of non ground goals, but
the authors do not aim at exploiting it to the degree we want.\todo{sucks}

Another work that is, surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.

@@

@@

\clearpage

\input{appendix.tex}

>>>>

\input{code/progress_heuristic}
%\noindent
%Note that \elpiIn{UV} is in \llambda as well as \elpiIn{T}. 
\vspace{-1em}
\section{Actual implementation in Elpi}\label{sec:implementation}

@@

\begin{elpicode}
  rule (N1 ~$\triangleright$~ G1 ?- link-eta (uvar X LX1) T1)   % match
    /  (N2 ~$\triangleright$~ G2 ?- link-eta (uvar X LX2) T2)   % remove
    |  (relocate LX1 LX2 T2 T2')               % condition
   <=> (N1 ~$\triangleright$~ G1 ?- T1 = T2').                  % new goal
\end{elpicode}

\noindent
The first directive matches a constraint
whose \lhs is a variable \elpiIn{X}; the second mathes and removes
a constraint on the same variable; the third
relocates \elpiIn{T2} and the latter crafs the new goal that unifies
\elpiIn{T1} with \elpiIn{T2'} under \elpiIn{G1} and the set of
heigenvariables \elpiIn{N1}.

@@

In the literature we could find a related encoding of the Calculus of
Constructions (CC)~\cite{felty93lics}. The goal of that work is to exhibit
a logic program performing proof checking for CC and hence relate the
proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
programs) with the one of CC. The encoding is hence tailored
toward a different goal, for example it utilizes three relations to represent the
equational theory of CC, and that choice alone makes things harder for us.
Section 6 contains a discussion about the use of the
unification procedure of the ML in presence of non ground goals, but
the authors do not aim at exploiting it to the degree we want.\todo{sucks}

Another work that is, somewhat surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.

@@

We considered mechanizing our results with Abella, especially in the early phases.
Unfortunately we could not. One reason
is that eliminating the cut operator, although possible, has a high cost in
verbosity.
% he same for the \elpiIn{name}
%\todo{is it sill described?} predicate.
Another problem is that the logic of Abella does not let one quantify over
predicates and combinators like \elpiIn{forall} are used pervasively in our
code.
\todo{finish}

@@

% \clearpage

% \input{appendix.tex}
